---
title: "Data 607 Final Project: Population Internet Access and Economic Growth"
authors: "Emily El Mouaquite, Long Lin, Pascal Hermann Kouogang Tafo, Ziaho Yu"
format: html
editor: visual
---

## Introduction

For this project, we want to investigate the relationship between internet access and economic opportunity through the following research question:

**Is higher internet access associated with stronger economic growth?**

To answer this question, we will integrate data from the online dataset [Share of the population using the Internet](https://ourworldindata.org/grapher/share-of-individuals-using-the-internet) (CSV file). Economic indicators, including GDP (USD) and GDP growth over time, will be retrieved through the [World Bank Indicators API](https://datahelpdesk.worldbank.org/knowledgebase/articles/889392-about-the-indicators-api-documentation), which provides JSON/XML data for all countries. Our motivation for doing this project is to see if, in a world where digital connectivity has become essential for education, commerce, and innovation, whether unequal access to the internet reinforces existing economic inequalities. By analyzing long‑term trends in internet usage alongside GDP growth, this project aims to quantify how strongly digital access correlates with national economic performance.

The workflow for this project will include: cleaning the internet access data from the CSV, and conducting statistical analyses to investigate the correlation between internet access and GDP growth over time. Possible hurdles that we might encounter could be misalignment between the CSV data and the data found in the API, missing data for some countries, and having to filter through the API data, which is very large.

## Fetching the GDP Data from the World Bank API

```{r}
library(httr2)
library(xml2)
library(tidyverse)

req <- request("https://api.worldbank.org/v2/country/all/indicator/NY.GDP.MKTP.KD.ZG") |>
  req_url_query(
    date = "1995:2023", 
    per_page = 32500
  )
resp <- req |> req_perform()
xml_data <- resp |> resp_body_xml()

records <- xml_find_all(xml_data, "//*[local-name()='data' and *[local-name()='indicator']]")

GDP_growth_df <- tibble(
  country     = records |> xml_find_first(".//*[local-name()='country']") |> xml_text(),
  country_iso = records |> xml_find_first(".//*[local-name()='countryiso3code']") |> xml_text(),
  date        = records |> xml_find_first(".//*[local-name()='date']") |> xml_integer(),
  value       = records |> xml_find_first(".//*[local-name()='value']") |> xml_double(),
  indicator   = records |> xml_find_first(".//*[local-name()='indicator']") |> xml_text()
)

print(GDP_growth_df)
```

```{r}
top_growth_df <- GDP_growth_df |>
  arrange(desc(value))

top_20_gdp_growth_df <- head(top_growth_df, 20)
glimpse(top_20_gdp_growth_df)
```

```{r}
bottom_growth_df <- GDP_growth_df |>
  arrange(value)

bottom_20_gdp_growth_df <- head(bottom_growth_df, 20)
glimpse(bottom_20_gdp_growth_df)
```

```{r}
library(httr2)
library(xml2)
library(tidyverse)

req <- request("https://api.worldbank.org/v2/country/all/indicator/SL.GDP.PCAP.EM.KD") |>
  req_url_query(
    date = "1995:2023", 
    per_page = 32500
  )
resp <- req |> req_perform()
xml_data <- resp |> resp_body_xml()

records <- xml_find_all(xml_data, "//*[local-name()='data' and *[local-name()='indicator']]")

GDP_per_person_df <- tibble(
  country     = records |> xml_find_first(".//*[local-name()='country']") |> xml_text(),
  country_iso = records |> xml_find_first(".//*[local-name()='countryiso3code']") |> xml_text(),
  date        = records |> xml_find_first(".//*[local-name()='date']") |> xml_integer(),
  value       = records |> xml_find_first(".//*[local-name()='value']") |> xml_double(),
  indicator   = records |> xml_find_first(".//*[local-name()='indicator']") |> xml_text()
)

print(GDP_per_person_df)
```

```{r}
top_GDP_per_person_df <- GDP_per_person_df |>
  arrange(desc(value))

head(top_GDP_per_person_df, 5)
```

## Internet Access CSV Cleaning/ Exploration

```{r}
library(tidyverse)
#read csv
internet_data <- read.csv("https://raw.githubusercontent.com/emilye5/607-final-project/refs/heads/main/share-of-individuals-using-the-internet.csv")
#clean dataset
internet_data_clean <- internet_data %>%
  #rename columns
  rename(
    country = Entity,
    country_iso = Code,
    date = Year,
    internet_usage_share = Share.of.the.population.using.the.Internet,
  ) %>%
  #drop rows with missing values 
  drop_na(internet_usage_share) %>%
  # keep only years after 1995 and before 2024
  filter(date >= 1995, date < 2024)
#take a look at the cleaned dataset
glimpse(internet_data_clean)
#check how many distinct countries are included in the data
n_distinct(internet_data_clean$country)
#boxplot showing the distribution of internet usage by year
internet_data_clean %>%
  ggplot(aes(x = factor(date), y = internet_usage_share)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Internet Usage Across Countries Over Time",
    x = "Year",
    y = "Internet Usage (%)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
final_data <- internet_data_clean |>
  left_join(GDP_per_person_df, by = c("country_iso", "date")) |>
  left_join(GDP_growth_df, by = c("country_iso", "date"))

final_data
```

```{r}
library(httr2)
library(xml2)
library(tidyverse)

req <- request("https://api.worldbank.org/v2/country/all/indicator/NY.GDP.MKTP.CD") |>
  req_url_query(
    date = "1995", 
    per_page = 32500
  )
resp <- req |> req_perform()
xml_data <- resp |> resp_body_xml()

records <- xml_find_all(xml_data, "//*[local-name()='data' and *[local-name()='indicator']]")

GDP_1995_df <- tibble(
  country     = records |> xml_find_first(".//*[local-name()='country']") |> xml_text(),
  country_iso = records |> xml_find_first(".//*[local-name()='countryiso3code']") |> xml_text(),
  date        = records |> xml_find_first(".//*[local-name()='date']") |> xml_integer(),
  value_1995       = records |> xml_find_first(".//*[local-name()='value']") |> xml_double(),
  indicator   = records |> xml_find_first(".//*[local-name()='indicator']") |> xml_text()
)

print(GDP_1995_df)

```

```{r}
GDP_1995_Remove_Regions_df <- GDP_1995_df[-(1:49), ]
print(GDP_1995_Remove_Regions_df)
```

```{r}
library(httr2)
library(xml2)
library(tidyverse)

req <- request("https://api.worldbank.org/v2/country/all/indicator/NY.GDP.MKTP.CD") |>
  req_url_query(
    date = "2023", 
    per_page = 32500
  )
resp <- req |> req_perform()
xml_data <- resp |> resp_body_xml()

records <- xml_find_all(xml_data, "//*[local-name()='data' and *[local-name()='indicator']]")

GDP_2023_df <- tibble(
  country     = records |> xml_find_first(".//*[local-name()='country']") |> xml_text(),
  country_iso = records |> xml_find_first(".//*[local-name()='countryiso3code']") |> xml_text(),
  date        = records |> xml_find_first(".//*[local-name()='date']") |> xml_integer(),
  value_2023       = records |> xml_find_first(".//*[local-name()='value']") |> xml_double(),
  indicator   = records |> xml_find_first(".//*[local-name()='indicator']") |> xml_text()
)

print(GDP_2023_df)
```

```{r}
GDP_2023_Remove_Regions_df <- GDP_2023_df[-(1:49), ]
print(GDP_2023_Remove_Regions_df)
```

```{r}
  Changed_GDP <- GDP_1995_Remove_Regions_df |>
  left_join(GDP_2023_Remove_Regions_df, by = c("country_iso"))
  print(Changed_GDP)
```

```{r}
  Changed_GDP_Average <- Changed_GDP |>
    mutate(
      change_in_gdp = value_2023 - value_1995,
      average_change = change_in_gdp / 28
    )
  print(Changed_GDP_Average)
```

```{r}
  top_avg_growth_df <- Changed_GDP_Average |>
    arrange(desc(average_change))
  
  top_20_avg_gdp_growth_df <- head(top_avg_growth_df, 20)
  top_20_avg_gdp_growth_df
```

```{r}
top_20_avg_gdp_growth_clean <- top_20_avg_gdp_growth_df %>%
  mutate(
    "Average Annual Change in USD (billions)" = average_change/1000000000
  ) %>%
  rename(Country = country.x) %>%
  select(-country.y, -indicator.x, -indicator.y, -value_1995, -value_2023, -change_in_gdp, -average_change, -date.x, -date.y)

top_20_avg_gdp_growth_clean
```

```{r}
 bottom_avg_growth_df <- Changed_GDP_Average |>
     arrange(average_change)
bottom_20_avg_gdp_growth_df <- head(bottom_avg_growth_df, 20)
bottom_20_avg_gdp_growth_df
```

```{r}
bottom_avg_gdp_growth_clean <- bottom_avg_growth_df %>%
  mutate(
    "Average Annual Change in USD (billions)" = average_change/1000000000
  ) %>%
  rename(Country = country.x) %>%
  select(-country.y, -indicator.x, -indicator.y, -value_1995, -value_2023, -change_in_gdp, -average_change, -date.x, -date.y)

bottom_avg_gdp_growth_clean

```
# Scatter plot: internet usage vs GDP growth (all countries, all years)

```{r}

correlation_data <- internet_data_clean |>
  inner_join(GDP_growth_df, by = c("country_iso", "date")) |>
  rename(country = country.x) |>  # resolve the .x/.y conflict after join
  drop_na(internet_usage_share, value)

country_correlations <- correlation_data |>
  group_by(country, country_iso) |>
  filter(n() >= 10) |>
  summarise(
    correlation = cor(internet_usage_share, value, method = "pearson"),
    n_years     = n(),
    .groups = "drop"
  ) |>
  arrange(desc(correlation))

# Top 10 positive correlations
cat("\n--- Top 10 Positive Correlations ---\n")
print(head(country_correlations, 10))

# Top 10 negative correlations
cat("\n--- Top 10 Negative Correlations ---\n")
print(tail(country_correlations, 10))

# Scatter plot

correlation_data |>
  ggplot(aes(x = internet_usage_share, y = value)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(
    title = "Internet Usage vs GDP Growth (1995–2023)",
    subtitle = paste("Overall Pearson r =", round(overall_cor, 3)),
    x = "Internet Usage Share (%)",
    y = "GDP Growth Rate (%)"
  ) +
  theme_minimal()

```



```{r, fig.width=18, fig.height=10}
selected_years <- c(1995, 2005, 2015, 2023)

world_years <- world[rep(seq_len(nrow(world)), times = length(selected_years)), ]
world_years$date <- rep(selected_years, each = nrow(world))

map_data_over_years <- world_years |>
  left_join(
    final_data |>
      filter(date %in% selected_years) |>
      select(country_iso, date, internet_usage_share, value.y),
    by = c("iso_a3" = "country_iso", "date" = "date")
  ) |>
  group_by(date) |>
  mutate(
    internet_group = ifelse(
      internet_usage_share >= median(internet_usage_share, na.rm = TRUE),
      "High Internet",
      "Low Internet"
    ),
    gdp_growth_group = ifelse(
      value.y >= median(value.y, na.rm = TRUE),
      "High GDP Growth",
      "Low GDP Growth"
    ),
    internet_gdp_group = ifelse(
      is.na(internet_usage_share) | is.na(value.y),
      NA,
      paste(internet_group, gdp_growth_group, sep = " + ")
    )
  ) |>
  ungroup()

tmap_mode("plot")

tmap_options(component.autoscale = FALSE)

tm_shape(map_data_over_years) +
  tm_polygons(
    "internet_gdp_group",
    title = "Internet Access and GDP Growth",
    colorNA = "brown",
    textNA = "No data"
  ) +
  tm_facets(
    by = "date",
    ncol = 2
  ) +
  tm_layout(
    main.title = "Internet Access and GDP Growth Over Time",
    main.title.size = 1.2,
    main.title.position = "center",
    panel.label.size = 1.1,
    legend.outside = TRUE,
    legend.outside.position = "right",
    legend.title.size = 0.9,
    legend.text.size = 0.8,
    inner.margins = c(0.01, 0.01, 0.01, 0.01),
    outer.margins = c(0.02, 0.02, 0.02, 0.02),
    frame = FALSE
  )
```


