---
title: "Data 607 Final Project: Population Internet Access and Economic Growth"
authors: "Emily El Mouaquite, Long Lin, Pascal Hermann Kouogang Tafo, Zihao Yu"
format: html
editor: visual
---

## Introduction

For this project, we want to investigate the relationship between internet access and economic opportunity through the following research question:

**Is higher internet access associated with stronger economic growth?**

To answer this question, we will integrate data from the online dataset [Share of the population using the Internet](https://ourworldindata.org/grapher/share-of-individuals-using-the-internet) (CSV file). Economic indicators, including GDP (USD) and GDP growth over time, will be retrieved through the [World Bank Indicators API](https://datahelpdesk.worldbank.org/knowledgebase/articles/889392-about-the-indicators-api-documentation), which provides JSON/XML data for all countries. Our motivation for doing this project is to see if, in a world where digital connectivity has become essential for education, commerce, and innovation, whether unequal access to the internet reinforces existing economic inequalities. By analyzing long‑term trends in internet usage alongside GDP growth, this project aims to quantify how strongly digital access correlates with national economic performance.

The workflow for this project will include: cleaning the internet access data from the CSV, and conducting statistical analyses to investigate the correlation between internet access and GDP growth over time. Possible hurdles that we might encounter could be misalignment between the CSV data and the data found in the API, missing data for some countries, and having to filter through the API data, which is very large.

### Packages and Libraries

```{r}
library(httr2)
library(xml2)
library(tidyverse)
library(tmap) #used for creating maps
library(sf) #used for tmap coordinates 
library(rnaturalearth) #provides map vectors for creating tmaps
```

### Fetching the GDP Data from the World Bank API

```{r}

req <- request("https://api.worldbank.org/v2/country/all/indicator/NY.GDP.MKTP.KD.ZG") |>
  req_url_query(
    date = "1995:2023", 
    per_page = 32500
  )
resp <- req |> req_perform()
xml_data <- resp |> resp_body_xml()

records <- xml_find_all(xml_data, "//*[local-name()='data' and *[local-name()='indicator']]")

GDP_growth_df <- tibble(
  country     = records |> xml_find_first(".//*[local-name()='country']") |> xml_text(),
  country_iso = records |> xml_find_first(".//*[local-name()='countryiso3code']") |> xml_text(),
  date        = records |> xml_find_first(".//*[local-name()='date']") |> xml_integer(),
  value       = records |> xml_find_first(".//*[local-name()='value']") |> xml_double(),
  indicator   = records |> xml_find_first(".//*[local-name()='indicator']") |> xml_text()
)

head(GDP_growth_df)
```

### Internet Access Data Cleaning/ Exploration

```{r}
#read csv
internet_data <- read.csv("https://raw.githubusercontent.com/emilye5/607-final-project/refs/heads/main/share-of-individuals-using-the-internet.csv")
#clean dataset
internet_data_clean <- internet_data %>%
  #rename columns
  rename(
    country = Entity,
    country_iso = Code,
    date = Year,
    internet_usage_share = Share.of.the.population.using.the.Internet,
  ) %>%
  #drop rows with missing values 
  drop_na(internet_usage_share) %>%
  # keep only years after 1995 and before 2024
  filter(date >= 1995, date < 2024)
#take a look at the cleaned dataset
glimpse(internet_data_clean)
#check how many distinct countries are included in the data
n_distinct(internet_data_clean$country)
```

### Distribution of Internet Usage Over time

```{r}
#boxplot showing the distribution of internet usage by year
internet_data_clean %>%
  ggplot(aes(x = factor(date), y = internet_usage_share)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Internet Usage Across Countries Over Time",
    x = "Year",
    y = "Internet Usage (%)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

In order to validate our internet usage data we created the above boxplot, which shows an increase in internet usage per year. The median half of the data increases in percentage of internet usage each year, with fewer outliers over time.

### Merging Internet & GDP Data

```{r}
final_data <- internet_data_clean |>
  left_join(GDP_growth_df, by = c("country_iso", "date"))

head(final_data)
```

### Fetching GDP Data from 1995 & 2023

1995:

```{r}
req <- request("https://api.worldbank.org/v2/country/all/indicator/NY.GDP.MKTP.CD") |>
  req_url_query(
    date = "1995", 
    per_page = 32500
  )
resp <- req |> req_perform()
xml_data <- resp |> resp_body_xml()

records <- xml_find_all(xml_data, "//*[local-name()='data' and *[local-name()='indicator']]")

GDP_1995_df <- tibble(
  country     = records |> xml_find_first(".//*[local-name()='country']") |> xml_text(),
  country_iso = records |> xml_find_first(".//*[local-name()='countryiso3code']") |> xml_text(),
  date        = records |> xml_find_first(".//*[local-name()='date']") |> xml_integer(),
  value_1995       = records |> xml_find_first(".//*[local-name()='value']") |> xml_double(),
  indicator   = records |> xml_find_first(".//*[local-name()='indicator']") |> xml_text()
)

print(GDP_1995_df)

```

```{r}
GDP_1995_Remove_Regions_df <- GDP_1995_df[-(1:49), ]
print(GDP_1995_Remove_Regions_df)
```

2023:

```{r}
req <- request("https://api.worldbank.org/v2/country/all/indicator/NY.GDP.MKTP.CD") |>
  req_url_query(
    date = "2023", 
    per_page = 32500
  )
resp <- req |> req_perform()
xml_data <- resp |> resp_body_xml()

records <- xml_find_all(xml_data, "//*[local-name()='data' and *[local-name()='indicator']]")

GDP_2023_df <- tibble(
  country     = records |> xml_find_first(".//*[local-name()='country']") |> xml_text(),
  country_iso = records |> xml_find_first(".//*[local-name()='countryiso3code']") |> xml_text(),
  date        = records |> xml_find_first(".//*[local-name()='date']") |> xml_integer(),
  value_2023       = records |> xml_find_first(".//*[local-name()='value']") |> xml_double(),
  indicator   = records |> xml_find_first(".//*[local-name()='indicator']") |> xml_text()
)

print(GDP_2023_df)
```

```{r}
GDP_2023_Remove_Regions_df <- GDP_2023_df[-(1:49), ]
print(GDP_2023_Remove_Regions_df)
```

```{r}
  Changed_GDP <- GDP_1995_Remove_Regions_df |>
  left_join(GDP_2023_Remove_Regions_df, by = c("country_iso"))
  print(Changed_GDP)
```

We removed the first 49 observations from the dataframe because the World Bank API includes world regions, which are not a part of our internet usage data.

### Calculating the Average Change in GDP by Country

```{r}
  Changed_GDP_Average <- Changed_GDP |>
    mutate(
      change_in_gdp = value_2023 - value_1995,
      average_change = change_in_gdp / 28
    )
  print(Changed_GDP_Average)
```

### Top 20 Countries Ranked by Average GDP Growth

```{r}
  top_avg_growth_df <- Changed_GDP_Average |>
    arrange(desc(average_change))
  
  top_20_avg_gdp_growth_df <- head(top_avg_growth_df, 20)
  top_20_avg_gdp_growth_df
```

```{r}
top_20_avg_gdp_growth_clean <- top_20_avg_gdp_growth_df %>%
  mutate(
    "Average Annual Change in USD (billions)" = average_change/1000000000
  ) %>%
  rename(Country = country.x) %>%
  #columns cleaned to remove date and indicator (always the same)
  select(-country.y, -indicator.x, -indicator.y, -value_1995, -value_2023, -change_in_gdp, -average_change, -date.x, -date.y)

top_20_avg_gdp_growth_clean
```

### Bottom 20 Countries Ranked by Average GDP Growth

```{r}
 bottom_avg_growth_df <- Changed_GDP_Average |>
     arrange(average_change)
bottom_20_avg_gdp_growth_df <- head(bottom_avg_growth_df, 20)
bottom_20_avg_gdp_growth_df
```

```{r}
bottom_avg_gdp_growth_clean <- bottom_avg_growth_df %>%
  mutate(
    "Average Annual Change in USD (billions)" = average_change/1000000000
  ) %>%
  #columns cleaned to remove date and indicator (always the same)
  rename(Country = country.x) %>%
  select(-country.y, -indicator.x, -indicator.y, -value_1995, -value_2023, -change_in_gdp, -average_change, -date.x, -date.y)

bottom_avg_gdp_growth_clean

```

### Calculating the Correlation Betweeen GDP and Internet Usage

```{r}
correlation_data <- internet_data_clean |>
  inner_join(GDP_growth_df, by = c("country_iso", "date")) |>
  rename(country = country.x) |>  # resolve the .x/.y conflict after join
  drop_na(internet_usage_share, value)

overall_cor <- cor(
  correlation_data$internet_usage_share,
  correlation_data$value,
  method = "pearson",
  use = "complete.obs"
)
print(overall_cor)

country_correlations <- correlation_data |>
  group_by(country, country_iso) |>
  filter(n() >= 10) |>
  summarise(
    correlation = cor(internet_usage_share, value, method = "pearson"),
    n_years     = n(),
    .groups = "drop"
  ) |>
  arrange(desc(correlation))
```

The Pearson Correlation Coefficient is -0.159. That indicates a very weak inverse linear relationship between Internet Usage and Annual GDP Growth. Moreover the effect size is small suggesting the relationship is not practically meaningful.

### Countries with the top 10 positive & top 10 negative correlations between internet usage and GDP

```{r}
# Top 10 positive correlations
cat("\n--- Top 10 Positive Correlations ---\n")
print(head(country_correlations, 10))

# Top 10 negative correlations
cat("\n--- Top 10 Negative Correlations ---\n")
print(tail(country_correlations, 10))
```

### Scatter plot: internet usage vs GDP growth (all countries, all years)

```{r}
# Scatter plot

correlation_data |>
  ggplot(aes(x = internet_usage_share, y = value)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(
    title = "Internet Usage vs GDP Growth (1995–2023)",
    subtitle = paste("Overall Pearson r =", round(overall_cor, 3)),
    x = "Internet Usage Share (%)",
    y = "GDP Growth Rate (%)"
  ) +
  theme_minimal()

```

To answer our main research question, we decided to generate a Scatter Plot and compute the Pearson Correlation to specifically quantify how strongly “Internet Usage” and the “ Annual GDP Growth” in all countries move together in a linear direction.

### Correlation Between Internet Usage and GDP by Year

```{r}
yearly_correlation <- final_data |>
  drop_na(internet_usage_share, value) |>
  group_by(date) |>
  summarize(
    correlation = cor(internet_usage_share, value, method = "pearson"),
    n_countries = n(),
    .groups = "drop"
  )

yearly_correlation

yearly_correlation |>
  ggplot(aes(x = date, y = correlation)) +
  geom_line(color = "steelblue") +
  geom_point(color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Yearly Correlation Between Internet Usage and GDP Growth",
    x = "Year",
    y = "Pearson Correlation"
  ) +
  theme_minimal()

```

The yearly correlation results show that the relationship between internet usage and GDP growth is not stable over time. Most years show weak or negative correlations, which suggests that higher internet usage is not consistently associated with higher GDP growth. This means that the association between internet usage and GDP growth is not consistent, and does not show a clear causal relationship.

### Mapping Internet Usage Over Time

```{r, fig.width=15, fig.height=9}

world <- ne_countries(scale = "medium", returnclass = "sf")

selected_years <- c(1995, 2005, 2015, 2023)

world_years <- world[rep(seq_len(nrow(world)), times = length(selected_years)), ]
world_years$date <- rep(selected_years, each = nrow(world))

map_internet_over_years <- world_years |>
  left_join(
    internet_data_clean |>
      filter(date %in% selected_years) |>
      select(country_iso, date, internet_usage_share),
    by = c("iso_a3" = "country_iso", "date" = "date")
  )

tmap_mode("plot")
tmap_options(component.autoscale = FALSE)

tm_shape(map_internet_over_years) +
  tm_polygons(
    "internet_usage_share",
    title = "Internet Usage (%)",
    style = "quantile",
    palette = "Blues",
    colorNA = "brown",
    textNA = "No data"
  ) +
  tm_facets(
    by = "date",
    ncol = 2
  ) +
  tm_layout(
    main.title = "Internet Usage by Country Over Time",
    main.title.size = 1.2,
    main.title.position = "center",
    panel.label.size = 1.1,
    legend.outside = TRUE,
    legend.outside.position = "right",
    legend.title.size = 0.9,
    legend.text.size = 0.8,
    inner.margins = c(0.01, 0.01, 0.01, 0.01),
    outer.margins = c(0.02, 0.02, 0.02, 0.02),
    frame = FALSE
  )
```

These maps show the global spread of internet usage from 1995 to 2023. Internet usage was limited in many countries in 1995, but by 2023, most regions show much higher levels of internet access.

### Mapping Internet Access and GDP Growth Over Time

```{r, fig.width=18, fig.height=10}
world <- ne_countries(scale = "medium", returnclass = "sf")

selected_years <- c(1995, 2005, 2015, 2023)

world_years <- world[rep(seq_len(nrow(world)), times = length(selected_years)), ]
world_years$date <- rep(selected_years, each = nrow(world))

map_data_over_years <- world_years |>
  left_join(
    final_data |>
      filter(date %in% selected_years) |>
      select(country_iso, date, internet_usage_share, value),
    by = c("iso_a3" = "country_iso", "date" = "date")
  ) |>
  group_by(date) |>
  mutate(
    internet_group = ifelse(
      internet_usage_share >= median(internet_usage_share, na.rm = TRUE),
      "High Internet",
      "Low Internet"
    ),
    gdp_growth_group = ifelse(
      value >= median(value, na.rm = TRUE),
      "High GDP Growth",
      "Low GDP Growth"
    ),
    internet_gdp_group = ifelse(
      is.na(internet_usage_share) | is.na(value),
      NA,
      paste(internet_group, gdp_growth_group, sep = " + ")
    )
  ) |>
  ungroup()

tmap_mode("plot")

tmap_options(component.autoscale = FALSE)

tm_shape(map_data_over_years) +
  tm_polygons(
    "internet_gdp_group",
    title = "Internet Access and GDP Growth",
    colorNA = "brown",
    textNA = "No data"
  ) +
  tm_facets(
    by = "date",
    ncol = 2
  ) +
  tm_layout(
    main.title = "Internet Access and GDP Growth Over Time",
    main.title.size = 1.2,
    main.title.position = "center",
    panel.label.size = 1.1,
    legend.outside = TRUE,
    legend.outside.position = "right",
    legend.title.size = 0.9,
    legend.text.size = 0.8,
    inner.margins = c(0.01, 0.01, 0.01, 0.01),
    outer.margins = c(0.02, 0.02, 0.02, 0.02),
    frame = FALSE
  )
```

These maps compare countries by whether they are above or below the yearly median for internet usage and GDP growth. It shows that the relationship between internet access and GDP growth varies between countries and years, rather than following one clear global pattern.

## Conclusion

Our findings suggest a weak inverse linear relationship between internet usage and GDP growth. The overall Pearson correlation is -0.159, which is close to 0 and indicates only a weak association. This means the data does not show a strong relationship between countries’ internet access and their GDP growth, meaning that internet usage, by itself, is not a significant enough parameter to determine the growth of a country's GDP. Therefore, this correlation analysis does not provide evidence that higher internet access is clearly associated with higher GDP growth.
